Augmentation of a Model Historical Harpsichord Mechanism for Haptic Enabled Interaction

### 2025-01-18 src/article.tex
fix refs

With the advent of digital musical instruments in the 20th Century there
has been a decoupling between interface and sound generation when
playing. Consequently, aspects of the sensory experience of playing have
been lost, namely the sensation of vibration and mechanical impedance
that a traditional musical instrument provides.

The project look at restoring tactile and haptic response of historical
harpsichords for an exhibition at the musical instrument Museo di San
Colombano\footnote{https://genusbononiae.it/en/san-colombano-tagliavini-collection/}
in Bologna. San Colombano is host to the Tagliavini Collection, a
collection of approximately 70 historical keyboard instruments in
working order \cite{Tagliavini2007} gifted to the museum by\\
from Luigi Ferdinando Tagliavini in 2010
\cite{SanColombano2010,Carlino2010}.

Some items among the Tagliavini collection are no longer in a playing
condition or are too fragile to continue to be available by museum
visitors. One of those fragile instruments is a 1547 Harpsichord in the
italian style by Alessandro Trasuntino
\cite{Wraight2024}\footnote{examples of similar instruments by
Trasuntino available online in the Royal College of Music Collection (1531:
\url{https://museumcollections.rcm.ac.uk/collection/Details/collect/58}) and the
Musée de la musique collection (1538:
\url{https://carmentis.kmkg-mrah.be:443/eMP/eMuseumPlus?service=ExternalInterface&module=collection&objectId=106088&viewType=detailView})
}, which was used as the basis for an interactive exhibition commisioned
by San Colombano.

Project was also designed to allow others to easily recreate and iterate
on the designs.

First a small cross-sectional model was used as a proof of concept. Once
a sensing system was decided upon, a larger 49-key scale model was
commissioned with and then built by luthier Roberto Livi. The models are
designed to provide only a tactile and haptic response for the player. A
sensor system was installed into the model and messages were sent to
computer that triggered sounds from a sample library which the player
listens to through headphones. The paper looks at the design decisions,
processes, and considerations to be made when undertaking a project of
this kind.

@@

This is particularly pertinent as it relates to conservation of musical
instruments, as they represent both historic objects, but also tools for
creating music. Historical musical instruments present an interesting
problem for cultural heritage as they are both a physical object but
also a tool for making music. This begs the question, can we use haptics
to preserve experiences of the past? The proposed project intends to
develop further the methodologies implemented in musical haptics
projects \cite{MusicalHaptics2018} such as \cite{Timmermans2020}.

@@

The project was carried out in collaboratoiun with San Colombano. The
results of the project would be presented as an interactive exhibition
for the public.

>>>>

With the advent of digital musical instruments in the 20th century there has
been a decoupling between interface and sound generation when playing.
Consequently, aspects of the sensory experience of playing have been lost,
namely the sensation of vibration and mechanical impedance that a traditional
musical instrument provides.

The project aims to restore the tactile and haptic response of historical
harpsichords for an exhibition at the Museo di San
Colombano\footnote{https://genusbononiae.it/en/san-colombano-tagliavini-collection/}
in Bologna. San Colombano is home to the Tagliavini Collection, a collection of
approximately 70 historical keyboard instruments in working order
\cite{Tagliavini2007} gifted to the museum by the Museum.\\
from Luigi Ferdinando Tagliavini in 2010 \cite{SanColombano2010,Carlino2010}.

Some items in the Tagliavini collection are no longer in playing condition or
are too fragile to continue to be available to museum visitors. One of those
fragile instruments is a 1547 Harpsichord in the Italian style by Alessandro
Trasuntino \cite{Wraight2024}\footnote{examples of similar instruments by
Trasuntino available online in the Royal College of Music Collection (1531:
\url{https://museumcollections.rcm.ac.uk/collection/Details/collect/58}) and the
Musée de la musique collection (1538:
\url{https://carmentis.kmkg-mrah.be:443/eMP/eMuseumPlus?service=ExternalInterface&module=collection&objectId=106088&viewType=detailView})
}, which was used as the basis for an interactive exhibition organised by San
Colombano.

The project was also open-sourced to allow for others to easily recreate and
iterate on the designs.

The "Haptic Key" project by Timmermans et al. \cite{Timmermans2020} began with
the model of a piano key. A 3-key cross-sectional model of a harpischord
mechanism \ref{fig:3key} was provided to the NEMUS project by San Colombano as a
test bed for sensors and as proof of concept for a larger scale interface. Each
key of the model has 2 sets of strings that are under tension but not at any
specific musical pitch. The strings are there to provide the same resistance to
the jack plectra as would be found on an instrument. The mechanism is in the
Italian style of the XVI-XIX century with 2 jacks per key and their
corresponding strings tuned in unison \footnote{"8'-8' È la dispozione normale
dei clavicembali italiani" \cite{Tagliavini1987}}.

A suitable sensing system was found to reliably detect when a jack had plucked a
string independently and then send a corresponding MIDI message to a connected
audio engine. A full scale 49-key scale model was commissioned with
and then built by luthier Roberto Livi in the same style as the 1547 Trasuntino
\ref{fig:teaser}. The sensor system was installed into the full-scale model and
then set up for exhibition at San Colombano. Visitors are invited to play on the
interface, which currently triggers samples on a connected sound engine. The
paper looks at the design decisions, processes, and considerations that need to
be made when undertaking a project of this kind.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/side_w-o_sensors.png} 
  \caption{3-key Model Harpsichord Mechanism by Graziano Bandini, Bologna 2023} 
  \Description{A 3-Key Model of a Harpsichord Mechanism created Graziano Bandini in Bologna. A side view of the model which is a cross-section of what is a common
mechanism for an Italian style harpsichord of the XVI century.} 
  \label{fig:3key}
\end{figure}

@@

This is particularly pertinent as it relates to the conservation of musical
instruments, as they represent both historic objects, but also tools for
creating music. Historical musical instruments present an interesting
problem for cultural heritage as they are both a physical object, but
also a tool for making music. This begs the question, can we use haptics
to preserve experiences of the past? The proposed project intends to
develop further the methodologies implemented in musical haptics
projects \cite{MusicalHaptics2018} such as \cite{Timmermans2020}.

@@

The project was carried out in collaboration with San Colombano. The
results of the project would be presented as an interactive exhibition
for the public.
